---
title: "Analysis of Mass Shootings in America"
author: "Teresa, Nick, Josh, Merzia"
subtitle: "MGSC 310, Fall 2019, Professor Hersh"
date: "12/12/2019"
output:
  keep_md: rmarkdown::github_document
---
  
# Environment Setup
```{r setup, echo=TRUE, eval=TRUE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)



# data manupulation
install.packages("tidyverse")
install.packages("magrittr")
install.packages("zoo")
install.packages("lubridate")
library(tidyverse)
library(magrittr)
library(zoo)
library(lubridate)

# plots
install.packages("ggplot2")
install.packages("gganimate")
install.packages("ggridges")
install.packages("hdm")
library(ggplot2)
library(gganimate)
library(ggridges)
library(hdm)

# models
install.packages("randomForest")
install.packages("glmnet")
install.packages("glmnetUtils")
library(randomForest)
library(glmnet)
library(glmnetUtils)

# diagnostics
install.packages("caret")
library(caret)
```

# Data Aquisition and Cleaning

## Loading the Datasets
```{r load data, eval=TRUE}
stanford_msa <- read.csv("../data/Stanford_MSA_Database.csv")
NYTimes_msa <- read.csv("../data/NYTimes_msa.csv")
```

## Data Cleaning
```{r data cleaning, warning=FALSE, eval=TRUE}
# aggregate similar factors
stanford_msa %<>% 
  mutate(Shooter.Race = fct_recode(Shooter.Race,
                                   "Asian American" = "Asian American/Some other race",
                                   "Black American or African American" = "Black American or African American/Unknown",
                                   "White American or European American" = "White American or European American/Some other Race",
                                   "Some other race" = "Some Other Race",
                                   "Some other race" = "Unknown"),
         Fate.of.Shooter.at.the.scene = fct_recode(Fate.of.Shooter.at.the.scene,
                                                   "Custody" = "Arrested",
                                                   "Deceased" = "Killed"),
         Fate.of.Shooter = fct_recode(Fate.of.Shooter,
                                      "Custody/Escaped" = "Custody / Escaped",
                                      "Custody/Escaped" = "FALSE"),
         Type.of.Gun...General = fct_recode(Type.of.Gun...General,
                                            "Multiple Guns" = "\nMultiple guns",
                                            "Multiple Guns" = "Multiple guns",
                                            "Multiple Guns" = "Multiple guns\n",
                                            "Handgun" = "handgun",
                                            "Handgun" = "9-mm"),
         Shooter.s.Cause.of.Death = fct_recode(Shooter.s.Cause.of.Death,
                                               "Suicide" = "Killed/Suicide",
                                               "Not Applicable" = "Not applicable",
                                               "Not Applicable" = "Not Apllicable"),
         Place.Type = fct_recode(Place.Type, "Park/Wilderness" = "Park/Wildness"),
         Targeted.Victim.s...General = fct_recode(Targeted.Victim.s...General,"General public" = "Social/General public"),
         Possible.Motive...General = fct_recode(Possible.Motive...General, 
                                                "Unknown" = "",
                                                "Unknown" = "Terminated/Denied/Reprimanded")) 

# correct column types
stanford_msa %<>% mutate(
  Title = as.character(Title),
  Location = as.character(Location),
  City = as.character(City),
  Shooter.Name = as.character(Shooter.Name),
  Shooter.Age.s. = toString(Shooter.Age.s.),
  Number.of.shooters = as.numeric(as.character(Number.of.shooters)),
  Average.Shooter.Age = as.numeric(as.character(Average.Shooter.Age)),
  Number.of.Shotguns = as.numeric(as.character(Number.of.Shotguns)),
  Number.of.Rifles = as.numeric(as.character(Number.of.Rifles)),
  Number.of.Handguns = as.numeric(as.character(Number.of.Handguns)),
  Number.of.Automatic.Guns = as.numeric(as.character(Number.of.Automatic.Guns)),
  Number.of.Semi.Automatic.Guns = as.numeric(as.character(Number.of.Semi.Automatic.Guns)),
  Total.Number.of.Guns = as.numeric(as.character(Total.Number.of.Guns)),
  Date = as.Date(Date, format="%m/%d/%Y")
)

# remove description columns
stanford_msa %<>% select(-c(CaseID, Description, Possible.Motive...Detailed, 
                            History.of.Mental.Illness...Detailed, Date...Detailed,
                            Targeted.Victim.s...Detailed, Type.of.Gun...Detailed, Notes,
                            Data.Source.1, Data.Source.2, Data.Source.3, Data.Source.4, 
                            Data.Source.5, Data.Source.6, Data.Source.7))

# filter cases with a high confidence depreciation value
stanford_msa %<>% filter(Depreciation == "1") # 1' indicates the case clearly fits the criteria for inclusion in the database

# replace NAs for numeric values with mean
is.numeric.NA <- function(x) return (is.numeric(x) & any(is.na(x)))
replace.NA.mean <- function(x) {
  avg <- round(mean(x, na.rm = T), 0)
  return (ifelse(is.na(x), avg, x))
}

stanford_msa %<>% mutate_if(is.numeric.NA, replace.NA.mean)

# correct column types
NYTimes_msa %<>% mutate(
  Publish.Date = as.Date(Publish.Date)
)
View(stanford_msa)
```

## Joining news data
```{r news, eval=TRUE}
dates <- unique(stanford_msa$Date)
article_countByDate <- data.frame()


for(i in 1:length(dates)) {
  curr_date <- dates[[i]]
  count <- nrow(
    NYTimes_msa %>% 
      filter(Publish.Date >= curr_date-30 & Publish.Date < curr_date)
    )
  
  article_countByDate %<>% bind_rows(data.frame(curr_date, count))
}

stanford_msa <- inner_join(stanford_msa, article_countByDate, by = c("Date" = "curr_date"))
stanford_msa %>% select(Date, count)
```

# Exploratory Plots

## Density ridge plot of Total Victims by Type of Gun
```{r graph, fig.height=5, fig.width=10, eval=TRUE}
ggplot(stanford_msa, aes(x = Total.Number.of.Victims, y = Type.of.Gun...General, fill = Type.of.Gun...General)) +
  geom_density_ridges(alpha = .5)
```

## Map of mass shootings
```{r map, eval=TRUE}
usa <- map_data('state')
ggplot() +
geom_path(data = usa, aes(x = long, y = lat, group = group)) +
geom_point(data = stanford_msa %>% filter(Longitude > -130), aes(x = Longitude, y = Latitude, size = Total.Number.of.Victims), color = "blue") + geom_density()
```

## Animated plots on Count & Total Victims over time
```{r anim, eval=TRUE}
stanford_msa %<>% mutate(year = year(Date), yearmon = as.yearmon(Date))

g <- ggplot(stanford_msa %>% filter(Date > "2000-01-01"), 
             aes(x = yearmon, y = count)) +
  geom_line(color= "blue") +
  transition_reveal(year)

g2 <- ggplot(stanford_msa %>% filter(Date > "2000-01-01"), 
             aes(x = yearmon, y = Total.Number.of.Victims, 
                 color=History.of.Mental.Illness...General)) +
  geom_line() +
  transition_reveal(year)

g
g2
```

# Analysis

## Select variables and training data
```{r subset}
msa_sub <- stanford_msa %>% select(Total.Number.of.Victims, Day.of.Week, Number.of.shooters,
                                   Average.Shooter.Age, Shooter.Sex, Shooter.Race,
                                   Number.of.Automatic.Guns, Number.of.Semi.Automatic.Guns, 
                                   Number.of.Shotguns, Fate.of.Shooter.at.the.scene, Fate.of.Shooter,
                                   Shooter.s.Cause.of.Death, Place.Type, Relationship.to.Incident.Location,
                                   Relationship.to.Incident.Location, Targeted.Victim.s...General, 
                                   Possible.Motive...General, History.of.Mental.Illness...General, 
                                   Military.Experience, Class, count)



train_idx <- sample(1:nrow(msa_sub), size = .80*nrow(msa_sub))
msa_train <- msa_sub %>% slice(train_idx)
msa_test <- msa_sub %>% slice(-train_idx)
```


## Random Forest
```{r rf}
rf_mod <- randomForest(Total.Number.of.Victims ~ ., data = msa_train,
                       type = regression,
                       mtry = 6,
                       ntree = 275,
                       importance = T)

# Helpful plots
plot(rf_mod)
varImpPlot(rf_mod)
randomForestExplainer::plot_min_depth_distribution(rf_mod)

diagnostics <- function(p, a) {
  print(RMSE(p, a))
  print(MAE(p, a))
  print(R2(p, a))
}
preds_train <- predict(rf_mod, newdata = msa_train, type = "response")
preds_test <- predict(rf_mod, newdata = msa_test, type = "response")

diagnostics(preds_train, msa_train$Total.Number.of.Victims)
diagnostics(preds_test, msa_test$Total.Number.of.Victims)
```

## Elastic Net
```{r elnet}


alpha_grid <- seq(0,1, len = 11)
alpha_grid

enet_fit <- cva.glmnet(Total.Number.of.Victims ~ .,
                       data = msa_train,
                       alpha = alpha_grid)
print(enet_fit)
summary(enet_fit)

minlossplot(enet_fit)

plot(enet_fit)

enet_preds_train <- predict(enet_fit, newdata = msa_train, alpha = 0, lambda = 5.97685)
enet_preds_train

enet_preds_test <- predict(enet_fit, newdata = msa_test, alpha = 0, lambda = 5.97685, type = 'response')
enet_preds_test

diagnostics(enet_preds_testm msa_test$Total.Number.of.Victims)
dignostics(enet_preds_train, msa_train$Total.Number.of.Victims)
```

#### Use ridge penalty
```{r ridge}
ridge_mod <- cv.glmnet(Total.Number.of.Victims ~ .,
                       data = msa_train,
                       alpha = 0)
summary(ridge_mod)


plot(ridge_mod)

ridge_mod$lambda.min

#training predictions
ridge_preds_train <- predict(ridge_mod, newdata = msa_train, s = "lambda.min", type = 'response')
ridge_preds_train



#testing predictions
ridge_preds_test <- predict(ridge_mod, newdata = msa_test, s = 'lambda.min', type = 'response')
ridge_preds_test

diagnostics(ridge_preds_train, msa_train$Total.Number.of.Victims)
diagnostics(ridge_preds_test, msa_test$Total.Number.of.Victims)
```

#### Get Ridge results
```{r ridge2}
lam_min_co <- as.matrix(coef(ridge_mod, s = "lambda.min"))
lam_min_co %>% subset(lam_min_co[,'1']< -1.5)
lam_min_co %>% subset(lam_min_co[,'1']> 3.5)
```

## K-means Clustering Model
```{r kmens}
# select clustering vars
msa_sub %<>% select_if(is.numeric) %>% select(-c(Number.of.Automatic.Guns, Number.of.Semi.Automatic.Guns, Number.of.Shotguns, count))

# silhouette method
fviz_nbclust(msa_sub, kmeans, method = "silhouette") +
  geom_vline(xintercept = 4, linetype = 2) +
  labs(subtitle = "Silhouette method")

# all methods
Nb_cl <- NbClust(msa_sub, diss = NULL, distance = "euclidean", 
                 min.nc = 2, max.nc = 21, method = "kmeans")

kmeans3 <- kmeans(msa_sub, centers = 3)

clusplot(msa_sub, kmeans3$cluster, color=TRUE, shade=FALSE, labels=5, lines=2)
fviz_cluster(kmeans3, data = msa_sub)
```






